Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2016-12-12 paper/mathematical_formulation_v4_cmame.tex
adding edits from karen

Consider two models with which parameters can be inferred: a high-fidelity model and a lower-fidelity model. Let the high-fidelity (HF) and low-fidelity (LF) models, and their corresponding variables, Lagrangians, and augmented Lagrangians, be distinguished by the subscripts $_{HF}$ and $_{LF}$, respectively. In particular, let $\Psi_{HF}$ and $\Psi_{LF}$ denote the stationary points of the high- and low- fidelity augmented Lagrangians $\mathcal{M}_{HF}$ and $\mathcal{M}_{LF}$, respectively. 
Then, the error in the Quantity of Interest $I$ is given by
%
\begin{multline}
\label{eq:semifinErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}(e^3)\textrm{.}
\end{multline}
%
where $\mathcal{R}$ is a remainder term that is third-order in the error $e=\Psi_{HF}-\Psi_{LF}$.
\end{proposition}
%
\begin{proof}
%
Observe that
%
\begin{equation}
\label{eq:MeqI}
\mathcal{M}(\Psi)=I(q,u),
\end{equation}
%
since taking variations of $\mathcal{M}$ with respect to the auxiliary variables gives that $\xi_\Psi$ is a stationary point of $\mathcal{L}$.

@@

Note that, as a stationary point, $\Psi_{HF}$ satisfies~\Cref{eq:supadjsys}. Now, let $\mathcal{Q}$ be an output defined by
%
\begin{equation}
\mathcal{Q}(\Phi)=\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Phi).
\label{eq:supadjout}
\end{equation}
%
Let the variational form
%
\begin{equation}
\mathscr{R}(\Psi_{HF})(\Phi)=0,\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:supadjsys}
\end{equation}
%
represent the equation $\mathcal{M}'_{HF,\Psi}(\Psi_{HF})(\Phi)=0$ in residual form, where $\Phi$ is a test function. Define the corresponding adjoint problem
%
\begin{equation}
\mathscr{R}_{\Psi}'(\Psi_{HF},\Phi)(\Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:superAdjEq}
\end{equation}
for the supplementary adjoint $\Lambda$, where $\mathscr{R}$ is defined in \Cref{eq:supadjsys}. The error in the output $\mathcal{Q}$ defined in \Cref{eq:supadjout} can then be expressed as a dual-weighted residual,
\begin{equation}
\label{eq:adjOutErr}
\mathcal M'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})=-\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda).
\end{equation}
%
Combining \Cref{eq:preadj} and \Cref{eq:adjOutErr}, we have,
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF})+\mathcal{R}(e^3)
\end{multline}
%
Other approaches can become feasible in certain scenarios; for example, if the high- and low-fidelity models only differ in the computational grid used, the difference $\Psi_{HF}-\Psi_{LF}$ can be approximated using interpolation or patch recovery methods~\cite{BecVex05}.

>>>>

Consider two models with which parameters can be inferred: a high-fidelity model and a lower-fidelity model. Let the high-fidelity (HF) and low-fidelity (LF) models, and their corresponding variables, Lagrangians, and augmented Lagrangians, be distinguished by the subscripts $_{HF}$ and $_{LF}$, respectively. In particular, let $\Psi_{HF}$ and $\Psi_{LF}$ denote the stationary points of the high- and low- fidelity augmented Lagrangians $\mathcal{M}_{HF}$ and $\mathcal{M}_{LF}$, respectively. 
Then, the error in the Quantity of Interest $I$ is given by
%
\begin{multline}
\label{eq:semifinErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}(e^3)\textrm{,}
\end{multline}
%
where $\mathcal{R}$ is a remainder term that is third-order in the error $e=\Psi_{HF}-\Psi_{LF}$.
\end{proposition}
%
\begin{proof}
%
Observe that
%
\begin{equation}
\label{eq:MeqI}
\mathcal{M}(\Psi)=I(q,u),
\end{equation}
%
since taking variations of $\mathcal{M}$ with respect to the auxiliary variables gives that $\xi_\Psi$ is a stationary point of $\mathcal{L}$.

@@

Note that, as a stationary point, $\Psi_{HF}$ satisfies~\Cref{eq:supadjsys}. Now, let $\mathcal{Q}$ be an output defined by
%
\begin{equation}
\mathcal{Q}(\Phi)=\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Phi).
\label{eq:supadjout}
\end{equation}
%
Let the variational form
%
\begin{equation}
\mathscr{R}(\Psi_{HF})(\Phi)=0,\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:supadjsys}
\end{equation}
%
represent the equation $\mathcal{M}'_{HF,\Psi}(\Psi_{HF})(\Phi)=0$ in residual form, where $\Phi$ is a test function. Define the corresponding adjoint problem
%
\begin{equation}
\mathscr{R}_{\Psi}'(\Psi_{HF},\Phi)(\Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:superAdjEq}
\end{equation}
for the supplementary adjoint $\Lambda$, where $\mathscr{R}$ is defined in \Cref{eq:supadjsys}. The error in the output $\mathcal{Q}$ defined in \Cref{eq:supadjout} can then be expressed as a dual-weighted residual,
\begin{equation}
\label{eq:adjOutErr}
\mathcal M'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})=-\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda).
\end{equation}
%
Combining \Cref{eq:preadj} and \Cref{eq:adjOutErr}, we have,
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF})+\mathcal{R}(e^3).
\end{multline}
%
Other approaches can become feasible in certain scenarios; for example, if the high- and low-fidelity models only differ in the computational grid used, the difference $\Psi_{HF}-\Psi_{LF}$ can be approximated using interpolation or patch recovery methods~\cite{BecVex05}.

### 2016-12-12 paper/numerical_experiments_v4_cmame.tex
adding edits from karen

\begin{figure}[htbp]
\centering
\captionsetup{justification=centering}
\subfloat[Locations of observations and QoI region $\Omega_I$][Locations of \\observations and \\QoI region $\Omega_I$]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_setup.png}
  \includegraphics[width=0.21\textwidth]{vs_qoi/vs_qoi_setup_sidetrim.png}
  \label{subfig:obsSetup}
}
\captionsetup{justification=centering}
\subfloat[MF$_0$ ($0\%$ HF)][MF$_0$ \\($0\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err0.png}
  \includegraphics[width=0.16\textwidth]{vs_qoi/vs_qoi_err0_nobar.png}
  \label{subfig:obsLF}
}
\captionsetup{justification=centering}
\subfloat[MF$_1$ ($\sim5\%$ HF)][MF$_1$ \\($\sim5\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err1.png}
  \includegraphics[width=0.16\textwidth]{vs_qoi/vs_qoi_err1_nobar.png}
}
\captionsetup{justification=centering}
\subfloat[MF$_2$ ($\sim10\%$ HF)][MF$_2$ \\($\sim10\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err2.png}
  \includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err2_barnorm.png}
  \label{subfig:obsMFlast}
}
  \caption{Effects of increasing the QoI region. Column \protect\subref{subfig:obsSetup}: configuration of observations (teal points) and QoI region (purple box). Columns \protect\subref{subfig:obsLF}--\protect\subref{subfig:obsMFlast}): the relative error estimate decompositions for different mixed-fidelity models, relative to the largest localized error contribution; note the locations of regions of relatively large error compared to the observation locations and QoI region.}
  \label{fig:qoiStudy}
\end{figure}

\Cref{fig:dataStudy} shows another set of cases considered, now with the same QoI region $\Omega_I$ but with increasing, nested sets of observations.
The error decomposition for the three cases is shown in \Cref{fig:dataStudy}. The bottom row is the same as that in \Cref{fig:qoiStudy}. Refinement appears to be consistently most important around the observation location closest to $x_1=0$ and the QoI region. However, as more observation locations are added, it becomes no longer necessarily true that refinement becomes less important around observation locations as their distance from the QoI region increases. In the second and third rows, one can see that after the areas around the QoI region and the two closest observation locations have been refined, the next area to be refined is not around the third closest observation location, but rather around one near the middle of the domain. This suggests that interactions between observation locations and the QoI may be non-intuitive, and in these cases a rigorous method for forming a mixed-fidelity model would be most helpful. % KW: I don't understand a lot of this discussion. You need to explain to me what you mean.

\begin{figure}[htbp]
\centering
\captionsetup{justification=centering}
\subfloat[Locations of observations and QoI region $\Omega_I$][Locations of \\observations and \\QoI region $\Omega_I$]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_setup.png}
  \includegraphics[width=0.21\textwidth]{vs_data/vs_data_setup_sidetrim.png}
  \label{subfig:obsSetup2}
}
\captionsetup{justification=centering}
\subfloat[MF$_0$ ($0\%$ HF)][MF$_0$ \\($0\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err0.png}
  \includegraphics[width=0.16\textwidth]{vs_data/vs_data_err0_nobar.png}
  \label{subfig:obsLF2}
}
\captionsetup{justification=centering}
\subfloat[MF$_1$ ($\sim5\%$ HF)][MF$_1$ \\($\sim5\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err1.png}
  \includegraphics[width=0.16\textwidth]{vs_data/vs_data_err1_nobar.png}
}
\captionsetup{justification=centering}
\subfloat[MF$_2$ ($\sim10\%$ HF)][MF$_2$ \\($\sim10\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err2.png}
  \includegraphics[width=0.23\textwidth]{vs_data/vs_data_err2_barnorm.png}
  \label{subfig:obsMFlast2}
}
  \caption{Effects of increasing the number of observations. Column \protect\subref{subfig:obsSetup2}: configuration of observations (teal points) and QoI region (purple box). Columns \protect\subref{subfig:obsLF2}--\protect\subref{subfig:obsMFlast2}): the relative error estimate decompositions for different mixed-fidelity models, relative to the largest localized error contribution; note the locations of regions of relatively large error compared to the observation locations and QoI region.}
  \label{fig:dataStudy}
\end{figure}

>>>>

\begin{figure}[htbp]
\centering
\captionsetup{justification=centering}
\subfloat[Locations of observations and QoI region $\Omega_I$][Locations of \\observations and \\QoI region $\Omega_I$]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_setup.png}
  \includegraphics[width=0.21\textwidth]{vs_qoi/vs_qoi_setup_sidetrim.png}
  \label{subfig:obsSetup}
}
\captionsetup{justification=centering}
\subfloat[MF$_0$ ($0\%$ HF)][MF$_0$ \\($0\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err0.png}
  \includegraphics[width=0.16\textwidth]{vs_qoi/vs_qoi_err0_nobar.png}
  \label{subfig:obsLF}
}
\captionsetup{justification=centering}
\subfloat[MF$_1$ ($\sim5\%$ HF)][MF$_1$ \\($\sim5\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err1.png}
  \includegraphics[width=0.16\textwidth]{vs_qoi/vs_qoi_err1_nobar.png}
}
\captionsetup{justification=centering}
\subfloat[MF$_2$ ($\sim10\%$ HF)][MF$_2$ \\($\sim10\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err2.png}
  \includegraphics[width=0.23\textwidth]{vs_qoi/vs_qoi_err2_barnorm.png}
  \label{subfig:obsMFlast}
}
  \caption{Effects of increasing the QoI region. Column \protect\subref{subfig:obsSetup}: configuration of observations (teal points) and QoI region (purple box). Columns \protect\subref{subfig:obsLF}--\protect\subref{subfig:obsMFlast}: the relative error estimate decompositions for different mixed-fidelity models, relative to the largest localized error contribution; note the locations of regions of relatively large error compared to the observation locations and QoI region.}
  \label{fig:qoiStudy}
\end{figure}

\Cref{fig:dataStudy} shows another set of cases considered, now with the same QoI region $\Omega_I$ but with increasing, nested sets of observations.
The error decomposition for the three cases is shown in \Cref{fig:dataStudy}. The bottom row is the same as that in \Cref{fig:qoiStudy}. Refinement appears to be consistently most important around the observation location closest to $x_1=0$ and the QoI region. However, as more observation locations are added, it becomes no longer necessarily true that refinement becomes less important around observation locations as their distance from the QoI region increases. In the second and third rows, one can see that after the areas around the QoI region and the two closest observation locations have been refined, the next area to be refined is not around the third closest observation location, but rather around one near the middle of the domain. This suggests that interactions between observation locations and the QoI may be non-intuitive, and in these cases a rigorous method for forming a mixed-fidelity model leads to well-supported modeling choices.

\begin{figure}[htbp]
\centering
\captionsetup{justification=centering}
\subfloat[Locations of observations and QoI region $\Omega_I$][Locations of \\observations and \\QoI region $\Omega_I$]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_setup.png}
  \includegraphics[width=0.21\textwidth]{vs_data/vs_data_setup_sidetrim.png}
  \label{subfig:obsSetup2}
}
\captionsetup{justification=centering}
\subfloat[MF$_0$ ($0\%$ HF)][MF$_0$ \\($0\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err0.png}
  \includegraphics[width=0.16\textwidth]{vs_data/vs_data_err0_nobar.png}
  \label{subfig:obsLF2}
}
\captionsetup{justification=centering}
\subfloat[MF$_1$ ($\sim5\%$ HF)][MF$_1$ \\($\sim5\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err1.png}
  \includegraphics[width=0.16\textwidth]{vs_data/vs_data_err1_nobar.png}
}
\captionsetup{justification=centering}
\subfloat[MF$_2$ ($\sim10\%$ HF)][MF$_2$ \\($\sim10\%$ HF)]{
  %\includegraphics[width=0.23\textwidth]{vs_data/vs_data_err2.png}
  \includegraphics[width=0.23\textwidth]{vs_data/vs_data_err2_barnorm.png}
  \label{subfig:obsMFlast2}
}
  \caption{Effects of increasing the number of observations. Column \protect\subref{subfig:obsSetup2}: configuration of observations (teal points) and QoI region (purple box). Columns \protect\subref{subfig:obsLF2}--\protect\subref{subfig:obsMFlast2}: the relative error estimate decompositions for different mixed-fidelity models, relative to the largest localized error contribution; note the locations of regions of relatively large error compared to the observation locations and QoI region.}
  \label{fig:dataStudy}
\end{figure}
